---
title: "County & Tract-level demographic data via TidyCensus"
author: "Shradha Dinesh"
date: "2024-06-03"
output: html_document
---

## Story memo

I will work on learning the TidyCensus R package to retrieve demographic data for all census tracts and counties nationally. Later, I will work with train route shapefiles, creating buffers and performing a spatial join to merge our demographic and geographic data. By the end of this week, I will have completed a high-level summary analysis of demographics for a few cities/towns along a specific coal (or train) route. 


Stories about particulate pollution in the industrial sector have described the air, earth, and our waterways. This would be the first comprehensive look at towns along coal transportation routes -- particularly prescient in the wake of major freight derailments, contamination events, and discussions about regulating the length and staff numbers aboard freight transport. 


The potential impact of this project could be increased investment (social and financial) for particulate monitoring sensors along transportation hubs and/or increased industrial regulation for coal transporters. The social impact would be increasing the public awareness of industrial air pollution or providing an explanation for the phenomenon in affected communities. 


This story would feature visual components like maps and other data graphics and photos and videos of coal trains and transportation routes. 


To complete this high level demographic analysis, I will need to begin working with train route shapefiles as a next step. 


Below, I have provided the code to produce dataframes with county and tract level demographic information, including race/ethnicity, poverty rate, and median income. These dataframes are stored in the `census_tract_stats` and `census_county_stats` variables. 


### Set up libraries

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
```

```{r}
# set api key for tidycensus
# census_api_key("bfa9ab659f49cf70e25323523f09670cbe4b98e5", install=TRUE)
# commenting out since already stored
```

### Instantiate global vars

```{r}
# instantiate global variable 
data(fips_codes)

## store to var
fips_codes <- fips_codes  
```

### Retrieve tract-level demographic data

```{r}

## Get list of states (Exclude non-states, except DC)
states <- fips_codes %>%
  select(state) %>%
  distinct() %>%
  head(51) %>%
  as_vector() 

# Get census tract data for all states
census_tract_stats <- get_acs(geography = "tract", variables = c( "B01001_001","B02001_002","B02001_003","B02001_004","B03001_003","B06012_002","B19013_001"), state=states, year = 2022) %>%
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  rename(
    total_pop = B01001_001,
    white_pop = B02001_002,
    black_pop = B02001_003,
    native_pop = B02001_004,
    hispanic_pop = B03001_003,
    poverty_pop = B06012_002,
    median_income = B19013_001
  ) %>%
  mutate(pct_white = round(white_pop/total_pop, 2) * 100,
         pct_nonwhite = 100 - round(white_pop/total_pop, 2) * 100,
         pct_black = round(black_pop/total_pop, 2) * 100,
         pct_native = round(native_pop/total_pop, 2) * 100,
         pct_hispanic = round(hispanic_pop/total_pop, 2) * 100,
         pct_poverty = round(poverty_pop/total_pop, 2) * 100
         ) %>%
  clean_names()

# Extract state and county FIPS codes from GEOID
census_tract_stats <- census_tract_stats %>%
  mutate(state_fips = substr(geoid, 1, 2),
         county_fips = substr(geoid, 3, 5))

# Join with fips_codes to get state and county names
fips_codes <- fips_codes %>%
  select(state_fips = state_code, county_fips = county_code, state, county)

census_tract_stats <- census_tract_stats %>%
  left_join(fips_codes, by = c("state_fips", "county_fips")) %>%
  select(geoid, state, county, starts_with("pct"), median_income)
```
### Retrieve county-level demographic data

```{r}
# Get county-level data for all states
census_county_stats <- get_acs(geography = "county", variables = c( "B01001_001","B02001_002","B02001_003","B02001_004","B03001_003","B06012_002","B19013_001"), year = 2022) %>%
  select(GEOID, NAME, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  rename(
    total_pop = B01001_001,
    white_pop = B02001_002,
    black_pop = B02001_003,
    native_pop = B02001_004,
    hispanic_pop = B03001_003,
    poverty_pop = B06012_002,
    median_income = B19013_001
  ) %>%
  mutate(pct_white = round(white_pop/total_pop, 2) * 100,
         pct_nonwhite = 100 - round(white_pop/total_pop, 2) * 100,
         pct_black = round(black_pop/total_pop, 2) * 100,
         pct_native = round(native_pop/total_pop, 2) * 100,
         pct_hispanic = round(hispanic_pop/total_pop, 2) * 100,
         pct_poverty = round(poverty_pop/total_pop, 2) * 100
         ) %>%
  clean_names() %>% 
  separate(name, into = c("county", "state"), sep = ", ")
```


